---
title: "Statistical Analysis: Valence Processing"
output: html_document
date: "2026-01-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE
)
```


# Preparation
In this section, we load the packages required for the subsequent analyses.

```{r}
# Data manipulation
library(tidyverse)
library(tidyr)
library(dplyr)

# Descriptive statistics
library(psych)

# Visualization
library(ggplot2)
library(corrplot)

# MIxed models
library(lme4)
library(lmerTest)

# Estimated marginal means
library (emmeans)
```


# Participant Information
Before analysing the experimental data, we examine whether participants' self-assessed English proficiency correlates with their LexTALE scores.
LexTALE is used as an objective measure of English vocabulary knowledge, while the self-assessment scores reflect participants’ perceived language abilities in four domains:

- speaking
- listening
- reading
- writing

## Correlation between LexTALE and Self-Assessment

The following data frame contains the LexTALE results and the self-assessment scores for each participant.

```{r}
# Dataframe of all self-assessment scores for each participant
participant_df <- data.frame(
  lex_tale_result = c(65,67.5,65,81.25,87.5,55,72.5,92.5,67.5,82.5,95,56.25,76.25,78.75,57.5,91.25),
  Speaking  = c(4,7,5,6,5,5,6,7,4,6,8,4,6,5,5,5),
  Listening = c(7,7,6,7,8,7,6,6,4,6,8,3,6,6,5,6),
  Reading   = c(5,7,7,6,8,6,6,5,5,7,8,3,6,7,5,7),
  Writing  = c(4,5,3,7,6,4,6,5,4,6,7,3,6,5,5,6)
)

# Correlations of each domain with LexTALE
cor(participant_df$lex_tale_result, participant_df$Speaking)
cor(participant_df$lex_tale_result, participant_df$Listening)
cor(participant_df$lex_tale_result, participant_df$Reading)
cor(participant_df$lex_tale_result, participant_df$Writing)

# Significance tests of the correlations
cor.test(participant_df$lex_tale_result, participant_df$Speaking)
cor.test(participant_df$lex_tale_result, participant_df$Listening)
cor.test(participant_df$lex_tale_result, participant_df$Reading)
cor.test(participant_df$lex_tale_result, participant_df$Writing)

# Matrix for nicer viewing
cor_matrix <- cor(participant_df)
cor_matrix
```


## Visualization of the Correlations
To visualise the relationship between LexTALE scores and self-assessed language skills, the data are transformed into long format and plotted separately for each skill.

```{r proficiency_correlation, cache=TRUE}
# Transform dataframe to long format
corr_df <- participant_df %>%
  pivot_longer(
    cols = Speaking:Writing,
    names_to = "skill",
    values_to = "score"
  )

# Plot
p <- ggplot(corr_df,
       aes(x = lex_tale_result, y = score)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "mediumpurple1", fill = "mediumpurple1") +
  facet_wrap(~skill) +
  theme_minimal() +
  labs(x = "LexTALE (%)", y = "Self-Assessment")

p


# Save
ggsave(
  "lextale_correlations.png",
  plot = p,
  width = 8,
  height = 6
)
```


# Experimental Data

## Loading the Dataset
The experimental dataset containing reaction times and accuracy values is loaded from a CSV file.

```{r}
descr_df <- read_csv("LDT_longlist_2026_01_09.csv", show_col_types = FALSE)
```


## Descriptive Statistics

Two main measures are analysed:

- **Accuracy** (whether the response was correct)
- **Reaction time (RT)** in milliseconds


### Initial Inspection of the Data
Before conducting statistical analyses, the dataset is filtered to include only experimental trials containing real words. This removes practice items and nonword trials.

```{r}
# Remove all practice items, remove all nonword trials
pre_filtered_df <- subset(descr_df, Item_type == "exp" & Condition == "word")
pre_filtered_df
```

### Participant-Level Summary Statistics

To identify potential outliers or problematic participants, we compute the mean accuracy and mean reaction time for each participant.

This allows us to detect:

- participants with extremely low accuracy (indicating misunderstanding of the task)
- participants with unusually slow responses

A preliminary plot is used to visualize potential outliers.

```{r participant_outliers, cache=TRUE}
# Overall overview by participant
describeBy(pre_filtered_df, group = pre_filtered_df$Participant)

# Summary: mean accuracy and RT by participant
means_by_participant <- pre_filtered_df %>%
  group_by(Participant) %>%
  summarise(
    mean_RT = mean(RT, na.rm = TRUE),
    mean_acc = mean(Accuracy, na.rm = TRUE)
  )
means_by_participant

# Plot (RT orange; accuracy green)
plot_outliers <- function(y_axis, filename, colour){
  p <- ggplot(data = means_by_participant, aes(x = "", y = {{ y_axis }})) +
  geom_boxplot(fill = colour, color = "black", alpha = 0.6) +
  labs(
    x = "",
    y = "Mean by Participant"
  ) +
  theme_minimal()

ggsave(
  filename,
  plot = last_plot(),
  width = 7,
  height = 5
)

p
}

plot_outliers(mean_RT, "average_RT_unfiltered.png", "darkolivegreen3")
plot_outliers(mean_acc, "average_acc_unfiltered.png", "sienna1")
```


One participant (ID 2400) shows extremely low accuracy (0.2, thus approximately 2%).
This strongly suggests that the participant did not perform the task correctly. Therefore, this participant is removed from the dataset.

We plot again to visualize what the data looks like now.

```{r removing_participants, cache=TRUE}
# Remove participant from the dataset
filtered_df <- subset(pre_filtered_df, !(Participant %in% c("2400")))
filtered_df

# Plot again:
means_by_participant <- filtered_df %>%
  group_by(Participant) %>%
  summarise(
    mean_RT = mean(RT, na.rm = TRUE),
    mean_acc = mean(Accuracy, na.rm = TRUE)
  )
means_by_participant

plot_outliers(mean_RT, "average_RT_filtered.png", "darkolivegreen3")
plot_outliers(mean_acc, "average_acc_filtered.png", "sienna1")
```


### Reaction Time Filtering

To remove implausible reaction times, a cutoff is applied.

Only trials with reaction times between **200 ms and 2000 ms** are retained.

This removes extremely fast responses (likely accidental key presses) and extremely slow responses that may reflect distractions or interruptions.

```{r reaction_time_filtering, cache=TRUE}
# Visualize the distribution of reaction times to see where to place the cutoff
p <- ggplot(data = filtered_df, aes(x = "", y = RT)) +
  geom_boxplot(fill = "darkolivegreen3", color = "black", alpha = 0.6) +
  labs(
    x = "",
    y = "Reaction Time"
  ) +
  theme_minimal()

#ggsave(
#  "RT_distribution.png",
#  plot = last_plot(),
#  width = 7,
#  height = 5
#)

p



# Apply cutoff
post_filtered_df <- filtered_df %>%
  filter(RT >= 200 & RT <= 2000)

# Plot again to visualize the new distribution
p <- ggplot(data = post_filtered_df, aes(x = "", y = RT)) +
  geom_boxplot(fill = "darkolivegreen3", color = "black", alpha = 0.6) +
  labs(
    x = "",
    y = "Reaction Time"
  ) +
  theme_minimal()

#ggsave(
#  "RT_outlier.png",
#  plot = last_plot(),
#  width = 7,
#  height = 5
#)

p

```

### Mean Accuracy and Reaction Time by Valence Condition
Next, mean accuracy and reaction time are computed for each emotional valence category:

- positive
- neutral
- negative


```{r mean_acc_RT_by_valence, cache=TRUE}
# Mean accuracy and reaction times by valence condition
means_by_valence <- post_filtered_df %>%
  group_by(Emotional_valence_cat) %>%
  summarise(
    mean_RT = mean(RT, na.rm = TRUE),
    mean_acc = mean(Accuracy, na.rm = TRUE)
  )
means_by_valence

# Plot mean accuracies (barplots)
plot_valence <- function(y_axis, filename, colour){
  p <- ggplot(data = means_by_valence, aes(x = Emotional_valence_cat, y = {{ y_axis }}, fill = Emotional_valence_cat)) +
    geom_col(fill = colour, width = 0.7) +
    labs(
      x = "Emotional Valence",
      y = "Mean Accuracy"
    ) +
    theme_minimal()

ggsave(
  filename,
  plot = last_plot(),
  width = 7,
  height = 5
)

p
}

plot_valence(mean_acc, "average_acc_per_valence_barplot.png", "sienna1")


# Plot mean RTs (boxplots)
plot_valence <- function(data, y_axis, filename, colour){

  p <- ggplot(data = data, 
              aes(x = Emotional_valence_cat, 
                  y = {{ y_axis }}, 
                  fill = Emotional_valence_cat)) +
    geom_boxplot(fill = colour, width = 0.7) +
    labs(
      x = "Emotional Valence",
      y = "RT"
    ) +
    theme_minimal()

  ggsave(
    filename,
    plot = p,
    width = 7,
    height = 5
  )

  # anzeigen
  p
}

plot_valence(post_filtered_df, RT, "RT_per_valence_boxplot.png", "darkolivegreen3")
```

# Inferential Statistics
To formally test the research questions, mixed-effects models are used.

Mixed-effects models are appropriate for experimental data because they
allow us to account for variability across both:

- participants
- stimuli


## Data Loading and Preparation

- Data loading: The full experimental dataset containing reaction times and accuracy values as well as participant data is loaded from a CSV file.
- Filtering: everything that was removed from the above dataset needs to be removed here as well (practice items, nonwords, certain participants)
- Variable coding: Before fitting the models, variables are converted to appropriate types (e.g., numeric or factor). The ordering of factor levels is also specified for interpretability.

```{r}
# Load dataset
inf_df <- read_csv("LDT_longlist_merged_2026_01_09.csv", show_col_types = FALSE, col_types = cols(
                          lex_tale_result = col_character()
                        )) %>%
  mutate(lex_tale_result = as.numeric(str_replace(lex_tale_result, ",", ".")))    # make sure LexTALE gets parsed right

# Filtering
prepped_df <- subset(inf_df, Item_type == "exp" & Condition == "word" & !(Participant %in% c("2400")) & RT >= 200 & RT <= 2000)

# Variable coding
prepped_df <- prepped_df %>%
  mutate(
    Accuracy = as.numeric(Accuracy),
    Participant = as.factor(Participant),
    Emotional_valence_cat = factor(Emotional_valence_cat, levels = c("neutral", "positive", "negative")),
    Length = as.numeric(Length),
    RT = as.numeric(RT),
    Stimulus = as.factor(Stimulus),
    lex_tale_result = as.numeric(lex_tale_result),
  )
```

### Standardizing Continuous Predictors

Continuous predictors are standardized (z-transformed). This improves model convergence and ensures that predictors are on comparable scales.

```{r}
prepped_df <- prepped_df %>%
  mutate(
    lex_tale_z = scale(lex_tale_result),
    Length_z   = scale(Length)
  )
```


## Accuracy Analysis

Accuracy is analysed using a **generalized linear mixed-effects model** with a binomial link function.

The model includes:

Fixed effects:

- LexTALE score (standardized)
- emotional valence
- word length

Random effects:

- random intercepts for participants
- random intercepts for stimuli
  
```{r ACC_mixed_model, cache=TRUE}
# Accuracy mixed model
acc_model <- glmer(Accuracy ~ lex_tale_z * Emotional_valence_cat + Length_z + (1| Participant) + (1| Stimulus), family = binomial, data = prepped_df, control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 1e5)))
summary(acc_model)

# Plot the model
# LexTALE (standardized)
lex_seq_z <- seq(min(prepped_df$lex_tale_z, na.rm = TRUE),
                 max(prepped_df$lex_tale_z, na.rm = TRUE),
                 by = 0.1)

# emmeans (standardized, length fixed to zero)
emm <- emmeans(
  acc_model,
  ~ lex_tale_z | Emotional_valence_cat,
  at = list(lex_tale_z = lex_seq_z, Length_z = 0),
  type = "response"
)

# Convert to dataframe
emm_df <- as.data.frame(emm)

# Backtransformation to original range (0–100%)
lex_mean <- mean(prepped_df$lex_tale_result)
lex_sd   <- sd(prepped_df$lex_tale_result)
emm_df$lex_tale_pct <- emm_df$lex_tale_z * lex_sd + lex_mean


# Plot
p <- ggplot(emm_df, aes(x = lex_tale_pct,
                   y = prob,
                   color = Emotional_valence_cat,
                   fill = Emotional_valence_cat)) +
  geom_line(linewidth = 1.2) +                     # Linie für geschätzte acc
  geom_ribbon(aes(ymin = asymp.LCL, ymax = asymp.UCL),
              alpha = 0.2, color = NA) +           # CI bänder
  labs(x = "LexTALE (%)",
       y = "Predicted Accuracy",
       color = "Emotional Valence",
       fill = "Emotional Valence") +
  scale_y_continuous(limits = c(0, 1)) +
  theme_minimal(base_size = 14)

ggsave(
  "acc_model.png",
  plot = last_plot(),
  width = 7,
  height = 5
)

p
```

## Reaction Time Analysis

Reaction times are analysed using a **linear mixed-effects model**.

Only trials with correct responses are included, as incorrect trials may reflect different cognitive processes.

The model includes:

Fixed effects:

- LexTALE score (standardized)
- emotional valence
- word length

Random effects:

- random intercepts for participants
- random intercepts for stimuli

```{r RT_mixed_model, cache=TRUE}
# Filter incorrect trials
RT_df <- subset(prepped_df, Accuracy == "1")

# RT mixed model
RT_model <- lmer(RT ~ lex_tale_z * Emotional_valence_cat + Length_z + (1| Participant) + (1| Stimulus), data = RT_df)
summary(RT_model)

# Plot the model
# LexTALE (standardized)
lex_seq_z <- seq(min(RT_df$lex_tale_z, na.rm = TRUE),
                 max(RT_df$lex_tale_z, na.rm = TRUE),
                 by = 0.1)

# emmeans (standardized, length fixed to zero)
emm <- emmeans(
  RT_model,
  ~ lex_tale_z | Emotional_valence_cat,
  at = list(lex_tale_z = lex_seq_z, Length_z = 0)
)

# Convert to dataframe
emm_df <- as.data.frame(emm)

# Backtransformation to original range (0-100%)
lex_mean <- mean(RT_df$lex_tale_result)
lex_sd   <- sd(RT_df$lex_tale_result)
emm_df$lex_tale_pct <- emm_df$lex_tale_z * lex_sd + lex_mean


# Plot
p <- ggplot(emm_df, aes(x = lex_tale_pct,
                   y = emmean,
                   color = Emotional_valence_cat,
                   fill = Emotional_valence_cat)) +
  geom_line(linewidth = 1.2) +
  geom_ribbon(aes(ymin = lower.CL, ymax = upper.CL),
              alpha = 0.2, color = NA) +
  labs(x = "LexTALE (%)",
       y = "Predicted RT (ms)",
       color = "Emotional Valence",
       fill = "Emotional Valence") +
  theme_minimal(base_size = 14)

ggsave(
  "RT_model.png",
  plot = last_plot(),
  width = 7,
  height = 5
)

p
```


# Interpretation

The results of the mixed-effects models are interpreted with respect to
the research questions:

1. Does English proficiency influence the **accuracy or speed** of lexical decisions?

2. Do **emotionally valenced words** (positive or negative) show a processing advantage compared to neutral words?

3. Does **emotional valence interact with English proficiency** in predicting reaction times or accuracy?

The answers to these questions are based on the inferential statistical results reported above.



# Additional Analyses
Post-hoc, we decided to additionally examine whether participants' overall performance in the lexical decision task correlates with their LexTALE scores.

```{r performance_LexTALE_correlation, cache=TRUE}
lex_by_participant <- prepped_df %>%
  group_by(Participant) %>%
  summarise(lex_tale_result = first(lex_tale_result))

means_by_participant$Participant <- as.character(means_by_participant$Participant)
lex_by_participant$Participant   <- as.character(lex_by_participant$Participant)

corr_df <- left_join(means_by_participant, lex_by_participant, by = "Participant")

# Accuracy

# Pearson
cor.test(corr_df$mean_acc,
         corr_df$lex_tale_result,
         method = "pearson")

# Plot
p <- ggplot(corr_df,
       aes(x = lex_tale_result, y = mean_acc)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "sienna1", fill = "sienna1") +
  theme_minimal() +
  labs(x = "LexTALE (%)", y = "Mean Accuracy")

ggsave(
  "acc_correlation.png",
  plot = last_plot(),
  width = 7,
  height = 5
)

p


# RT

#Pearson
cor.test(corr_df$mean_RT,
         corr_df$lex_tale_result,
         method = "pearson")

# Plot
p <- ggplot(corr_df,
       aes(x = lex_tale_result, y = mean_RT)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, color = "darkolivegreen3", fill = "darkolivegreen3") +
  theme_minimal() +
  labs(x = "LexTALE (%)", y = "Mean RT")

ggsave(
  "RT_correlation.png",
  plot = last_plot(),
  width = 7,
  height = 5
)

p
```
